---
title: "Global validation: strain ranks across diseases"
date: "2026-01-20"
---

# PURPOSE

analysis to support manuscript, specifically the statement:

Across nearly all diseases evaluated, Strain Recommender identified strains predicted to more closely resemble the disease state than the strains previously used and curated in the literature (Refs). Specifically, in over 88% of diseases, at least one strain with more extreme predicted vulnerability was identified relative to the strain reported in the MGD. Notably, 185 inbred strains (28.2%) were predicted to resemble disease states more frequently than C57BL/6J, the most commonly background for modeling human disease 

# METHODS

## Overview
This analysis performs the following key steps:
1. Read in filtered result table 
2. For each strain, calculate strain ranks based on vulnerability predictions
3. Compare strain ranks to observed strain ranks for each disease/geneset
4. Rank strains across diseases by frequency of 'vulnerable' classification
5. Determine where C57BL/6J ranks relative to other strains

## Load Required Packages


```{r}

# load packages
library(tidyverse)
library(data.table)
library(lme4)
library(tableone)
library(kableExtra)

```


## Set Output Directory

```{r}
outdir <- "/projects/chesler-lab/jcpg/imputation/robyn/data/retrospective/"
```


## Data Loading and Initial Processing

Read in the final filtered dataset containing retrospective analysis results.


```{r}

final_results <- fread("data/supplementary_table3.csv.gz", data.table = FALSE)
length(unique(final_results$gs_id))

# Check the number of unique values in non-numeric variables
# This helps understand the scope of the dataset
nvars <- c("fprA", "fprB", "distance_pred", "sensitivity", "max_distance")
apply(final_results[, -which(colnames(final_results) %in% nvars)], 2, function(x) length(unique(x)))

```
## Dataset Scope Assessment

Determine the number of strains and their inclusion status after filtering.


```{r}
s <- unique(c(final_results$strainA, final_results$strainB))
sinc <- unique(c(final_results$strainA[which(final_results$included_final==1)], final_results$strainB[which(final_results$included_final==1)]))
cat(length(s), "strains used in the analysis.\n", length(sinc), "included in the final analysis after filtering on fprA <= 0.05.\n")
```

## Gene Set Analysis Setup

Identify gene sets and their inclusion status for downstream analysis.


```{r}
# Path to strain recommender results
sr_results <- '/projects/chesler-lab/jcpg/retrospective/kleina/strain_recommender/results/mESC/gs_set1/'

# Get unique gene set IDs that passed filtering
gsids <- unique(final_results$gs_id[which(final_results$included_final == 1)])

# Create dataframe tracking all gene sets and their inclusion status
gs_all <- data.frame(gs_id = unique(final_results$gs_id))
gs_all$included <- as.numeric(gs_all$gs_id %in% gsids)

# Summary of gene set inclusion
cat("Gene set inclusion summary:\n")
print(table(gs_all$included, dnn = "Included"))
cat("Total:", nrow(gs_all), "eligible genesets (prior to filtering for strainA fpr <= 0.05)\n")
cat(sum(gs_all$included), "included,", sum(!gs_all$included), "not included\n")
```

## Strain Recommender Results Assembly

Process individual strain recommender result files for each gene set, applying FPR filtering.


```{r}
# Initialize empty dataframe to store all strain results
strain_results <- data.frame()

fns <- list.files(sr_results, full.names = TRUE)

# Process each gene set individually
for (gs in gs_all$gs_id) {
  
  # Find the appropriate filename for this gene set
  fn <- fns[grep(paste0(gs, "_.*"), fns)]
  
  # Handle cases with multiple or no matching filenames
  condition <- length(fn) > 1
  if (condition) {
    cat("Check filename specification for:", gs, ".\nThere are", length(fn), "filenames. Choosing 'pos1'...\n")
    fn <- fn[grep("pos1", fn)]
  } else if (length(fn) == 0) {
    cat("Check filename specification for:", gs, ".\nThere are", length(fn), "filenames. Skipping...\n")
  }
  
  # Read strain recommender results and filter for sex="both" and fpr <= 0.05
  sr <- fread(fn, data.table=FALSE) %>% 
    filter(sex == "both" & fpr <= 0.05)
  
  # If no results meet fpr <= 0.05 threshold, relax criteria
  if (nrow(sr) == 0) {
    # relax fpr
    sr <- fread(fn, data.table=FALSE) %>% 
      filter(sex == "both")
    
    # Find best positive and negative correlations (lowest FPR)
    fpr_pos <- sr$fpr[which.min(sr$fpr[which(sr$corr > 0)])]
    fpr_neg <- sr$fpr[which.min(sr$fpr[which(sr$corr < 0)])]
    
    # Keep results with FPR <= max of best positive or negative
    sr <- sr %>% 
      filter(fpr <= max(fpr_pos, fpr_neg))
  }
  
  # Process and standardize the results
  sr <- sr %>% 
    select(-sex) %>% 
    mutate(direction = ifelse(corr >= 0, 1, -1),
           gs = gs)
  
  # Append to main results dataframe
  strain_results <- rbind(strain_results, sr)
  
}

# Clean up and add gene set inclusion information
strain_results <- strain_results %>% 
  mutate(gs_id = gs) %>% 
  select(-gs) %>% 
  left_join(., gs_all, join_by(gs_id))

cat("Processed", length(unique(strain_results$gs_id)), "gene sets\n")
cat("Total strain-disease associations:", nrow(strain_results), "\n")
```

# ANALYSIS 1: Support Statement on Disease Coverage

**Target Statement:** "In over 88% of diseases, at least one strain with more extreme predicted vulnerability was identified relative to the strain reported in the MGD."

## Identify Best Performing Strains

For each gene set, identify the strain with the lowest FPR (best performance) in the appropriate direction.

```{r}
diseases <- unique(final_results$disease_name)
# add in rare
# rare <- final_results %>% 
#   filter(rare == 1) %>% 
#   select(disease_name) %>% 
#   distinct %>% t() %>% as.vector(., "character")
```

```{r}
# look at genesets included in the analysis.
# if score_direction != 'neg1',
#   annotate final results with the lowest fpr across all 'vulnerable' strains
# if score_direction == 'neg1', 
#   annotate final results with the lowest fpr across all 'resistant' strains
# 
# Initialize columns to store best strain information
final_results$lowest_fpr <- NA
final_results$best_strain <- ""
gs_problematic <- NULL

# Process each gene set to find best performing strains
for (gs in gs_all$gs_id) {
  
  # Identify records where score_direction == "neg1" (negative correlation expected)
  neg <- which(final_results$score_direction == "neg1" & final_results$gs_id == gs)
  
  # Identify records where score_direction != "neg1" (positive correlation expected) 
  pos <- which(final_results$score_direction != "neg1" & final_results$gs_id == gs)
  
  ## Handle negative correlation cases (score_direction == "neg1")
  condition <- length(neg) > 0
  if (condition) {
    
    # Find best FPR among strains with negative correlation (direction = -1)
    best_fpr <- strain_results %>% 
      filter(gs_id == gs & direction == -1) %>% 
      select(fpr) %>% 
      min(.)
    
    final_results$lowest_fpr[neg] <- best_fpr
    
    # Get strain(s) with best FPR
    best_strains <- strain_results %>% 
      filter(gs_id == gs & direction == -1 & fpr == best_fpr) %>% 
      select(strain) %>% t() %>% as.vector(., "character")
    
    final_results$best_strain[neg] <- best_strains[1]
    
    # If strainA is among the best strains, annotate it as the best strain
    condition <- final_results$strainA[neg] %in% best_strains
    if (any(condition)) {
      meets_condition <- neg[which(condition==TRUE)]
      final_results$best_strain[meets_condition] <- final_results$strainA[meets_condition]
    } 
    
  }
  
  ## Handle positive correlation cases (score_direction != "neg1")
  condition <- length(pos) > 0
  if (condition) {
    
    # Filter for positive correlations
    pos_sr <- strain_results %>% 
      filter(gs_id == gs & direction == 1)
    
    # Check if any positive correlations exist
    if (nrow(pos_sr) > 0) {
      best_fpr <-  pos_sr %>% 
        select(fpr) %>% 
        min(.)
      
      final_results$lowest_fpr[pos] <- best_fpr
      
      # Get best performing strains
      best_strains <- strain_results %>% 
        filter(gs_id == gs & direction == 1 & fpr == best_fpr) %>% 
        select(strain) %>% t() %>% as.vector(., "character")
      
      final_results$best_strain[pos] <- best_strains[1]
      
      # Check if strainA is among best performers. If so annotate strainA as the best strain.
      condition <- final_results$strainA[pos] %in% best_strains
      if (any(condition)) {
        meets_condition <- pos[which(condition==TRUE)]
        final_results$best_strain[meets_condition] <- final_results$strainA[meets_condition]
      } 
      
    } else {
      # No positive correlations found - use largest negative correlation
      gs_problematic <- c(gs_problematic, gs)
      
      # Find best available correlation (largest negative)
      best_fpr <- strain_results %>% 
        filter(gs_id == gs) %>% 
        select(corr, fpr) %>% 
        arrange(-corr, fpr) %>% 
        summarise(min(fpr[which.max(corr)])) %>% 
        as.vector(., "numeric")
      
      final_results$lowest_fpr[pos] <- best_fpr
      
      best_strains <- strain_results %>% 
        filter(gs_id == gs & fpr == best_fpr) %>% 
        select(strain) %>% t() %>% as.vector(., "character")
      
      final_results$best_strain[pos] <- best_strains[1]
      
      # Check if strainA is among best performers. If so annotate strainA as the best strain.
      condition <- final_results$strainA[pos] %in% best_strains
      if (any(condition)) {
        meets_condition <- pos[which(condition==TRUE)]
        final_results$best_strain[meets_condition] <- final_results$strainA[meets_condition]
      } 
    }
  }
}

# Report any problematic gene sets
if (length(gs_problematic) > 0) {
  cat("Problematic gene sets (no positive correlations):", length(gs_problematic), "\n")
}
```

## Calculate Disease Coverage Statistics

Determine what percentage of diseases have better-performing strains than those reported in MGD.


```{r}
# Calculate percentage where best strain differs from reported strain (strainA)
mu <- final_results %>% 
  select(gs_id, strainA, fprA, score_direction, disease_name, lowest_fpr, best_strain) %>% 
  distinct() %>% 
  summarise(mu = mean(strainA != best_strain)) %>% as.vector(., "numeric")

cat("Specifically, in over", round(mu*100),"% of diseases, at least one strain with more extreme predicted vulnerability was identified relative to the strain reported in the MGD.\n")
```


# ANALYSIS 2: Strain Ranking Analysis

**Target Statement:** "185 inbred strains (28.2%) were predicted to resemble disease states more frequently than C57BL/6J, the most commonly background for modeling human disease"

## Create Disease-Gene Set Mapping

Establish the relationship between diseases and gene sets for proper vulnerability assessment.

## Build Comprehensive Strain Vulnerability Table

Create a table showing which strains are vulnerable for each disease-gene set combination.


```{r}

disease_gs_score_direction <- final_results %>% 
  select(gs_id, disease_name, score_direction) %>% 
  mutate(disease = disease_name) %>% 
  select(-disease_name) %>% 
  distinct()

strain_table <- data.frame()
disease_table <- data.frame()

# Process each gene set to identify vulnerable/resistant strains
for (gs in gs_all$gs_id) {
  sr <- strain_results %>% 
    filter(gs_id == gs & fpr <= 0.05)
  
  # Only proceed if at least one strain met the FPR threshold
  if (nrow(sr) > 0) {
    vulnerable <- sr %>% filter(corr >= 0)
    resistant <- sr %>% filter(corr < 0)
    strains <- c(vulnerable$strain, resistant$strain)
    
    # Get associated diseases for this gene set
    diseases <- unique(final_results$disease_name[which(final_results$gs_id == gs)])
    
    # Create vulnerability indicator (1 = vulnerable, 0 = resistant)
    if (length(vulnerable) > 0) {
      is_vulnerable <- as.numeric(strains %in% vulnerable$strain)
    } else {
      is_vulnerable <- rep(0, length(strains))
    }
    
    # Create gene set table
    gs_table <- data.frame(gs_id = gs, strain = strains, vulnerable = is_vulnerable)
    
    # Replicate for each associated disease
    disease_vector <- rep(diseases, each = nrow(gs_table))
    disease_table <- rbind(disease_table, 
                           data.frame(disease = disease_vector, bind_rows(replicate(length(diseases), gs_table, simplify = FALSE))))
  }
}

# Adjust vulnerability based on score direction
# If score_direction == "neg1", flip the vulnerability interpretation
disease_table <- disease_table %>% 
  left_join(., disease_gs_score_direction, join_by(gs_id, disease)) %>% 
  mutate(vulnerable = ifelse(score_direction == "neg1", 1 - vulnerable, vulnerable))

# Save comprehensive disease-geneset-strain vulnerability table
# fwrite(disease_table, paste0(outdir, "retrospective_manuscript_disease_gs_strain_vulnerabilities_2026-01-23.csv.gz"), compress = "gzip")

cat("Created vulnerability table with", nrow(disease_table), "records\n")
cat("Covering", length(unique(disease_table$disease)), "diseases and", 
    length(unique(disease_table$strain)), "strains\n")
```

## Summarize Strain Performance

Create strain-level summary showing frequency of vulnerability across diseases.

```{r}
# Create disease-level strain vulnerability summary
strain_table <- disease_table %>% 
  select(disease, strain, vulnerable) %>% 
  distinct()
  
# Handle any inconsistencies (same strain-disease pair with different vulnerability scores)
inconsistent <- strain_table %>% 
  group_by(disease, strain) %>% 
  tally() 

# Resolve inconsistencies: if strain was vulnerable for disease using any geneset, mark as vulnerable
strain_table <- strain_table %>% 
  group_by(disease, strain) %>% 
  left_join(., inconsistent, join_by(disease, strain)) %>% 
  mutate(vulnerable = ifelse(n > 1, 1, vulnerable)) %>% 
  select(-n) %>% 
  distinct()

# Create strain-level summary statistics
strain_counts <- strain_table %>% 
  group_by(strain) %>% 
  summarise(extreme = n(), vulnerable=sum(vulnerable)) %>% 
  mutate(extreme_rank = rank(-extreme), 
         vulnerable_rank = rank(-vulnerable))

cat("Generated summary for", nrow(strain_counts), "strains\n")
```

## Save Summary Tables

Export the processed tables for further analysis and manuscript preparation.

```{r}
fwrite(strain_table, paste0(outdir, "retrospective_manuscript_disease_strain_vulnerabilities_2026-01-23.csv.gz"), compress="gzip")
fwrite(strain_counts, paste0(outdir, "retrospective_manuscript_strain_vulnerabilities_2026-01-23.csv.gz"), compress="gzip")
```

## C57BL/6J Ranking Analysis

Determine how C57BL/6J ranks relative to other strains in disease vulnerability frequency.

```{r}
# Get C57BL/6J performance metrics
b6_ranks <- strain_counts %>% 
  filter(strain == "C57BL/6J")

# Report results for diseases where strains are vulnerable
cat("Notably,", length(which(strain_counts$vulnerable_rank < b6_ranks$vulnerable_rank)), "inbred strains", round(mean(strain_counts$vulnerable_rank < b6_ranks$vulnerable_rank)*100,1), "% were predicted to resemble disease states more frequently than C57BL/6J, the most commonly background for modeling human disease.\n\n")

# Report results for diseases where strains are vulnerable or resistant
cat("Notably,", length(which(strain_counts$extreme_rank < b6_ranks$extreme_rank)), "inbred strains", round(mean(strain_counts$extreme_rank < b6_ranks$extreme_rank)*100,1), "% were predicted to be extreme across diseases more frequently than C57BL/6J, the most commonly background for modeling human disease")

```

```{r}
# Top 10 most vulnerable strains
cat("\nTop 10 strains by vulnerability frequency:\n")
top_vulnerable <- strain_counts %>% 
  arrange(vulnerable_rank) %>% 
  head(10) %>% 
  select(strain, vulnerable, extreme, vulnerable_rank)
print(top_vulnerable)
```



# Overall sensitivity

```{r}
final_results <- fread("data/supplementary_table3.csv.gz", data.table = FALSE)
```

```{r}
final_results %>% 
  filter(included_final == 1) %>% 
  summarise(x = sum(correct), n = n())
```

```{r}
library(binom)
#result <- binom.confint(x = 38351, n = 45089, methods = "wilson")
result <- binom.confint(x = 11197, n = 13132, methods = "wilson")
print(round(result[4:6]*100, 1))
```
## rare
```{r}
final_results %>% 
  filter(included_final == 1 & rare == 1) %>% 
  summarise(x = sum(correct), n = n())
```

```{r}
#result <- binom.confint(x = 19563, n = 24055, methods = "wilson")
result <- binom.confint(x = 5815, n = 7172, methods = "wilson")
print(round(result[4:6]*100, 1))
```

```{r}
final_results %>% 
  filter(included_final == 1) %>% 
  select(jnum_id) %>% 
  distinct() %>% 
  tally()
```

