---
title: "Disease sensitivity and sensitivity analysis (FPR threshold) across other tissue types"
---

Purpose: The following code reads in the global validation study results to assess sensitivity mean sensitivity when setting fpr cutoffs in 0.05 increments.

```{r}

# load necessary packages
library(tidyverse)
library(data.table) 
library(R.utils)
library(binom)
library(patchwork)

```

Source functions

unit tests for functions in unit_tests/

```{r}
setwd('/projects/chesler-lab/jcpg/retrospective/kleina/jcpg/recommender_manuscript/global_validation')
source("retrospective_function_process_data.R")
source("retrospective_function_get_sensitivity.R")
source("retrospective_function_check_exclusions.R")
```

Read in the results table before fprA filtering. This contains 129 gene sets.
```{r}
setwd('/projects/chesler-lab/jcpg/retrospective/kleina/jcpg/recommender_manuscript/global_validation')
mesc_results <- fread("data/supplementary_table3.csv.gz", data.table = FALSE)
length(unique(mesc_results$gs_id))
```

Read in the other tissue results tables before fprA filtering. These contain the same 129 gene sets as the mESC result table (`results`).
function: read in all processed tissue type results
```{r}

read_tissue_results <- function(dir_path = "/projects/chesler-lab/jcpg/retrospective/kleina/retrospective_analysis/results") {
  files <- list.files(dir_path, pattern = "_129_full_res_prefprfilter_corrected_final_dups_removed_2026-04-02.csv.gz", full.names = TRUE)
  names(files) <- gsub("_129_full_res_prefprfilter_corrected_final_dups_removed_2026-04-02.csv.gz", "", basename(files))
  files <- files[names(files) != "mESC"]
  
  res_list <- lapply(files, \(f) {
  message("Reading: ", basename(f))
  as_tibble(data.table::fread(f))
})
  
  # assign to global env as tissue_res
  iwalk(res_list, ~assign(paste0(.y, "_results"), .x, envir = .GlobalEnv))
  
  invisible(res_list)
}

```

```{r}

tissue_types <- c("mesc", "heart", "liver", "retina", "rpe", "bone", 
                   "pfc_saline", "skeletal", "striatum_saline", 
                   "adipose", "striatum", "islet", "hippocampus", "maxilla")

```

rare disease list accessed via Orphanet on Oct 25, 2024.  
```{r}

rare <- fread("/projects/chesler-lab/jcpg/retrospective/kleina/jcpg/recommender_manuscript/global_validation/data/List_of_rare_diseases_in_alphabetical_order.csv", data.table = FALSE)
#rare <- fread("data/List_of_rare_diseases_in_alphabetical_order.csv", data.table = FALSE)

```

read in each of the tissue result files, and when doing so, get an idea of the fprA distribution via a simple histogram, summary stats.
```{r}

res_list <- read_tissue_results()

plots <- imap(res_list, \(res, tissue) {
  message("Tissue type: ", tissue)
  print(summary(res$fprA))
  
  ggplot(data = res) +
    geom_histogram(aes(fprA), bins = 20) +
    theme_bw() +
    ggtitle(tissue)
})

```
look at the plots
```{r}

plots

```

calculate mean sensivitivty for each tissue
```{r}
# unique comparison variables
single <- c("gs_id", "strainA", "strainB", "MP_ID", "jnum_id", "disease_name")
# imputed transcriptome
#tissue <- "mESC"
thresholds <- c(0.05, 0.10, 0.15, 0.20)
 
```

```{r}

run_thresholds <- function(tissue, thresholds) {
  res <- get(paste0(tissue, "_results"))
  
  n_diseases <- NULL
  thresholded_results <- data.frame()
  
  for (threshold in thresholds) {
    main <- process_data(df_all = res, rare = rare, thresh = threshold, tissue = 'tissue')
    disease_sensitivities <- main[[2]]
    disease_sensitivities$thresh = threshold
    disease_sensitivities$tissue = tissue
    n_diseases <- c(n_diseases, nrow(disease_sensitivities))
    thresholded_results <- rbind(thresholded_results, disease_sensitivities)
  }
  
  names(n_diseases) <- thresholds
  message(tissue, ":")
  print(n_diseases)
  
  list(n_diseases = n_diseases, thresholded_results = thresholded_results)
}

```

run `thresholded_results()` across all tissues
```{r}

all_thresh_results <- lapply(tissue_types, \(t) {
  message("Processing: ", t)
  run_thresholds(t, thresholds)
})
names(all_thresh_results) <- tissue_types

combined_thresh_res <- bind_rows(lapply(all_thresh_results, \(x) x$thresholded_results))

```

# Calculate mean sensitivity and 95% CI for thresholds using threshold results

```{r}

calc_mean_sensitivity <- function(thresholded_results, thresholds) {
  
  mean_sensitivity <- data.frame()
  for (threshold in thresholds) {
    subset <- thresholded_results %>% 
      filter(thresh == threshold)
    n <- nrow(subset)
    p <- mean(subset$sensitivity) 
    sd <- sd(subset$sensitivity)
    s <- 1.96 * (sd / sqrt(n))
    mean_sensitivity <- rbind(mean_sensitivity,
                              data.frame(thresh = threshold, mean = round(p, 3),
                                         lower = round(p - s, 3), upper = round(p + s, 3)))
  }
  mean_sensitivity
}

```

```{r}

all_mean_sensitivity <- lapply(tissue_types, \(t) {
  tr <- all_thresh_results[[t]]$thresholded_results
  ms <- calc_mean_sensitivity(tr, thresholds)
  ms$tissue <- t
  ms
})

combined_mean_sensitivity <- bind_rows(all_mean_sensitivity)

combined_mean_sensitivity <- combined_mean_sensitivity %>% select(tissue, thresh, mean, lower, upper)

```

write reuslts to files
```{r}

fwrite(combined_thresh_res, file = '/projects/chesler-lab/jcpg/retrospective/kleina/jcpg/recommender_manuscript/global_validation/data/disease_sensitivity_by_thresh_2026-04-02.csv.gz', compress = 'gzip')
fwrite(combined_mean_sensitivity, file = '/projects/chesler-lab/jcpg/retrospective/kleina/jcpg/recommender_manuscript/global_validation/data/mean_sensitivity_other_tissues_by_thresh_2026-04-02.csv.gz', compress = 'gzip')

```


