---
title: "Sensitivity analysis: non-redundant comparisons across diseases"
---

# PURPOSE

Assess sensitivity on non-redundant comparisons introduced by the ontology mapping. 

# METHODS

A non-redundant subset of diseases was constructed by identifying diseases sharing identical sets of comparisons and retaining a single representative per unique comparison set. The first listed disease was used as the representative, yielding 2,019 diseases.

# CODE

Load packages

```{r, message=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(binom)
```

Source functions

```{r}
source("retrospective_function_check_exclusions.R")
source("retrospective_function_get_sensitivity.R")
source("retrospective_function_process_data.R")
```

Read in data

```{r}
df <- fread("data/supplementary_table3.csv.gz", data.table = FALSE)
```

# Distinct comparisons in final analysis: distinct sets of MP, jnum, strain comparison, geneset

choose the first-listed disease as the representative disease for this unique set of comparisons

```{r, message=FALSE}
disease_subset <- df %>% 
  group_by(disease_name) %>% 
  select(MP_ID, jnum_id, strainA, strainB, gs_id) %>% 
  distinct() %>% 
  summarise(set = sort(paste(MP_ID, jnum_id, strainA, strainB, gs_id, collapse = ","))) %>% 
  ungroup() %>% 
  group_by(set) %>% 
  summarise(n_diseases = n(), 
            diseases = paste(disease_name, collapse = ","),
            first_disease = first(disease_name)) %>%
  arrange(desc(n_diseases))

cat("There are", nrow(disease_subset), "representative diseases with a unique set of comparisons.\n")
```

# save 

```{r}
# df <- df %>% 
#   mutate(nonredundant_comparison = as.numeric(disease_name %in% disease_subset$first_disease)) 
# fwrite(df, "data/supplementary_table3.csv.gz", compress="gzip")
```


# Calculate disease-specific sensitivities on the non-redundant set of comparisons 


```{r}
# unique comparison variables
single <- c("gs_id", "strainA", "strainB", "MP_ID", "jnum_id", "disease_name")
# FPR threshold
thresh <- 0.05
# imputed transcriptome
tissue <- "mESC"
```

Rare disease list accessed via Orphanet on Oct 25, 2024.  
```{r}
rare <- fread("data/List_of_rare_diseases_in_alphabetical_order.csv", data.table = FALSE)
```

Run results through `process_data()` function. Write each part of result to object.

```{r}
df_subset <- df %>% 
  filter(disease_name %in% disease_subset$first_disease)

main <- process_data(df_all = df_subset, rare = rare)
results_filtered <- main[[1]]
disease_sensitivities <- main[[2]]
```

```{r}
round(summary(disease_sensitivities$sensitivity)*100,1)
```

```{r}
n <- length(disease_sensitivities$sensitivity)
p <- mean(disease_sensitivities$sensitivity)
sd <- sd(disease_sensitivities$sensitivity)
s <- 1.96*(sd/sqrt(n))
#95% CI
cat("The mean sensitivity =", round(p,3), "(95% CI", round(p - s,3), ",", round(p+s,3), ")\n") 
```

## Group diseases by sensitivity. Print number of diseases and corresponding percentages per sensitivity category.

```{r}
# count number of diseases
n_d_tested <- nrow(disease_sensitivities)

# determine number of diseases that fall into each sensitivity category
n_ge90 <- length(which(disease_sensitivities$sensitivity >= 0.9))
n_mod <- length(which(disease_sensitivities$sensitivity >= 0.7 & disease_sensitivities$sensitivity < 0.9))
n5070 <- length(which(disease_sensitivities$sensitivity > 0.5 & disease_sensitivities$sensitivity < 0.7))
nle50 <- length(which(disease_sensitivities$sensitivity <= 0.5))

# print number of diseases per sensitivity category
n_ge90
n_mod
n5070
nle50

# print percentage of diseases per sensitivity category
n_ge90 / n_d_tested * 100
n_mod/ n_d_tested *100
n5070/ n_d_tested *100
nle50/ n_d_tested *100
```
