---
title: "Sensitivity analysis: FPR threshold"
---

Purpose: The following code reads in the global validation study results to assess mean sensitivity when setting fpr cutoffs in 0.05 increments.

```{r}

# load necessary packages
library(tidyverse)
library(data.table) 
library(R.utils)
library(binom)
library(patchwork)

```

Source functions

unit tests for functions in unit_tests/

```{r}
source("retrospective_function_process_data.R")
source("retrospective_function_get_sensitivity.R")
source("retrospective_function_check_exclusions.R")
```

Read in the results table before fprA filtering. This contains 129 gene sets.
```{r}
results <- fread("data/supplementary_table3.csv.gz", data.table = FALSE)
length(unique(results$gs_id))
```

Get an idea of the fprA distribution via a simple histogram, summary stats.
```{r}

ggplot(data = results) + geom_histogram(aes(fprA), bins= 20) + theme_bw()

summary(results$fprA)

```

 Rare disease list accessed via Orphanet on Oct 25, 2024.  
```{r}

rare <- fread("data/List_of_rare_diseases_in_alphabetical_order.csv", data.table = FALSE)

```



```{r}
# unique comparison variables
single <- c("gs_id", "strainA", "strainB", "MP_ID", "jnum_id", "disease_name")
# imputed transcriptome
tissue <- "mESC"
```

```{r}
thresholds <- c(0.05, 0.10, 0.15, 0.20)
n_diseases <- NULL
thresholded_results <- data.frame()
for (threshold in thresholds) {
  main <- process_data(df_all = results, rare = rare, thresh = threshold)
  disease_sensitivities <- main[[2]]
  disease_sensitivities$thresh = threshold
  n_diseases <- c(n_diseases, nrow(disease_sensitivities))
  thresholded_results <- rbind(thresholded_results, disease_sensitivities)
}
```

# Calculate mean sensitivity and 95% CI for thresholds

```{r}
thresholds <- c(0.05, 0.10, 0.15, 0.20)
mean_sensitivity <- data.frame()
for (threshold in thresholds) {
  subset <- thresholded_results %>% 
    filter(thresh == threshold)
  n <- nrow(subset)
  p <- mean(subset$sensitivity) 
  sd <- sd(subset$sensitivity)
  s <- 1.96*(sd/sqrt(n))
  mean_sensitivity <- rbind(mean_sensitivity,
                            data.frame(thresh = threshold, mean=round(p,3),
                                       lower=round(p-s, 3), upper=round(p+s,3)))
}

mean_sensitivity
```
```{r}
names(n_diseases) <- thresholds
print(n_diseases)
```

