---
title: "Results: disease sensitivity calculations"
---

Purpose: The following code reads in the final filtered result set and calculates the disease sensitivity distribution across all diseases, as well as across the subset of rare diseases.

Data will be filtered to fprA <= 0.05 (as indicated by the `included_final` variable) in rlb's `process_data()` function.

```{r}

# load necessary packages
library(tidyverse)
library(data.table) 
library(R.utils)
library(binom)

```

Source functions

unit tests for functions in unit_tests/

```{r}
source("retrospective_function_process_data.R")
source("retrospective_function_get_sensitivity.R")
source("retrospective_function_check_exclusions.R")
```


Read in the results
```{r}

results  <- fread('data/supplementary_table3.csv.gz', sep = ',', data.table = FALSE)

```

Rare disease list accessed via Orphanet on Oct 25, 2024.  
```{r}

rare <- fread("data/List_of_rare_diseases_in_alphabetical_order.csv", data.table = FALSE)

```


```{r}
# unique comparison variables
single <- c("gs_id", "strainA", "strainB", "MP_ID", "jnum_id", "disease_name")
# FPR threshold
thresh <- 0.05
# imputed transcriptome
tissue <- "mESC"
```


Run results through `process_data()` function. Write each part of result to object.
```{r}

main <- process_data(df_all = results, rare = rare, thresh = 0.05)
results_filtered <- main[[1]]
disease_sensitivities <- main[[2]]

```

Disease-specific performance overview

```{r}
cat("Number of diseases:", nrow(disease_sensitivities), "\n")
cat("Summary of disease-specific sensitivities:\n")
print(summary(disease_sensitivities$sensitivity))
cat("\n")

n <- length(disease_sensitivities$sensitivity)
p <- mean(disease_sensitivities$sensitivity)
sd <- sd(disease_sensitivities$sensitivity)
s <- 1.96*(sd/sqrt(n))
cat("Mean sensitivity =", round(p,3), paste0(" (95% CI ", round(p - s,3), ",", round(p+s,3), ")."), "\n")

cat("\n\n")
cat("Number of rare diseases:", sum(disease_sensitivities$rare), "\n") 
cat("Proportion of rare diseases:", sum(disease_sensitivities$rare)/ nrow(disease_sensitivities), "\n")
cat("Summary of rare disease-specific sensitivities:\n")
print(summary(disease_sensitivities$sensitivity[which(disease_sensitivities$rare==1)]))
cat("\n")

is_rare <- which(disease_sensitivities$rare == 1)
n <- length(disease_sensitivities$sensitivity[is_rare])
p <- mean(disease_sensitivities$sensitivity[is_rare])
sd <- sd(disease_sensitivities$sensitivity[is_rare])
s <- 1.96*(sd/sqrt(n))
cat("Mean sensitivity of rare diseases =", round(p,3), paste0(" (95% CI ", round(p - s,3), ",", round(p+s,3), ")."), "\n")
```

Group diseases by sensitivity. Print number of diseases and corresponding percentages per sensitivity category.

```{r}

# count number of diseases
n_d_tested <- nrow(disease_sensitivities)

# determine number of diseases that fall into each sensitivity category
n_ge90 <- length(which(disease_sensitivities$sensitivity >= 0.9))
n_mod <- length(which(disease_sensitivities$sensitivity >= 0.7 & disease_sensitivities$sensitivity < 0.9))
n50 <- length(which(disease_sensitivities$sensitivity > 0.5 & disease_sensitivities$sensitivity < 0.7))
nl50 <- length(which(disease_sensitivities$sensitivity <= 0.5))

# print number of diseases per sensitivity category
n_ge90
n_mod
n50
nl50

# print percentage of diseases per sensitivity category
n_ge90 / n_d_tested * 100
n_mod/ n_d_tested *100
n50/ n_d_tested *100
nl50/ n_d_tested *100

```

Create sensitivity distribution pie chart for all diseases.
```{r}
# Pie Chart with Percentages
slices <- c( n_ge90, n_mod, n50, nl50)

lbls <- c("High sensitivity: ", "Moderately high sensitivity: ", "Moderate sensitivity: ", "Low sensitivity: ")
n <- slices
pct <- round(slices/sum(slices)*100)
lbls <- rep("",4)
pie(slices,labels = lbls, col=c("#225ea8", "#41b6c4", "#a1dab4", "#f7f7f7"),
   main="All diseases")

svg("data/results_pie_all.svg", bg = "transparent")
pie(slices,labels = lbls, col=c("#225ea8", "#41b6c4", "#a1dab4", "#f7f7f7"))
dev.off()

```

Now do the same things for the rare disease subset.
```{r}

# count number of rare diseases
r_d_tested <- length(which(disease_sensitivities$rare==1))

# determine number of rare diseases that fall into each sensitivity category
r_ge90 <- length(which(disease_sensitivities$sensitivity[which(disease_sensitivities$rare==1)] >= 0.9))
r_mod <- length(which(disease_sensitivities$sensitivity[which(disease_sensitivities$rare==1)] >= 0.7 & disease_sensitivities$sensitivity[which(disease_sensitivities$rare==1)] < 0.9))
r50 <- length(which(disease_sensitivities$sensitivity[which(disease_sensitivities$rare==1)] >= 0.5 & disease_sensitivities$sensitivity[which(disease_sensitivities$rare==1)] < 0.7))
rl50 <- length(which(disease_sensitivities$sensitivity[which(disease_sensitivities$rare==1)] < 0.5))

# print number of rare diseases per sensitivity category
r_ge90
r_mod
r50
rl50

# print percentage of rare diseases per sensitivity category
r_ge90 / r_d_tested*100
r_mod / r_d_tested*100
r50 / r_d_tested*100
rl50 / r_d_tested*100

```

Create sensitivity distribution pie chart for rare disease subset.
```{r}
# Pie Chart with Percentages
slices <- c( r_ge90, r_mod, r50, rl50)

lbls <- c("High sensitivity: ", "Moderately high sensitivity: ", "Moderate sensitivity: ", "Low sensitivity: ")
n <- slices
pct <- round(slices/sum(slices)*100)
lbls <- rep("",4)
pie(slices, labels=lbls, col=c("#225ea8", "#41b6c4", "#a1dab4", "#f7f7f7"))

svg("data/results_pie_rare.svg", bg = "transparent")
pie(slices,labels = lbls, col=c("#225ea8", "#41b6c4", "#a1dab4", "#f7f7f7"))
dev.off()
```



## Details about the final result set. 

Number of MP terms, studies, gene sets, unique strain pairs, unique comparisons, and unique strain names.

Number of MP terms.
```{r}

length(table(results_filtered$MP_ID))

```

Number of studies.
```{r}
length(table(results_filtered$jnum_id))

```

Number of gene sets.
```{r}
length(table(results_filtered$gs_id))

```

Number unique comparisons

```{r}
nrow(unique(results_filtered[which(results_filtered$included==1), c('gs_id', 'strainA', 'strainB', 'jnum_id', 'MP_ID')]))
```

Number unique strain comparisons

```{r}
nrow(unique(results_filtered[which(results_filtered$included==1), c('strainA', 'strainB', 'jnum_id', 'MP_ID')]))
```

Number of unique strain pairs.
```{r}

nrow(unique(results_filtered[, c('strainA', 'strainB')]))

```

Number of unique strains.
```{r}

length(unique(c(results_filtered$strainA, results_filtered$strainB)))

```

