---
title: "CTD Analysis"
date: "14-August-2025"
output: html_notebook
---

```{r}

# run before starting
# setwd("/projects/chesler-lab/jcpg/CTD_2025") 

.libPaths(c('/projects/chesler-lab/chesler-R01-2021/kleina/R/rocker-rstudio/4.3', '/projects/chesler-lab/chesler-R01-2021/kleina/R/rocker-rstudio/4.4', .libPaths()))
library(tidyverse)
library(data.table)
library(readxl)
library(e1071)

```

read in the data
```{r}

# full df
data_full <- fread(file = '/projects/chesler-lab/jcpg/CTD_2025/CTD_data_15July2025.csv', sep = ',', data.table = FALSE)
# filtered df- 7 mice with non-matching jr strains to CC036 or CC027 removed
data_filtered <- fread(file = '/projects/chesler-lab/jcpg/CTD_2025/CTD_data_filtered_13Aug2025.csv', sep = ',', data.table = FALSE)

```

set up variable lists for echo and ekg variables being used for models
```{r}

ekg_vars<- c('tpeak_tend_interval_msec', 't_amplitude_mv', 'qrs_interval_msec', 'jt_interval_msec', 'rr_interval_msec', 'p_duration_msec', 'st_amplitude_mv') # 7 variables
echo_vars<- c('asc_ao', 'linear2_aortic_annulus', 'linear3_aortic_sinus', 'linear4_sinotabular_junction', 'ef_sax_m_mode', 'fs_sax_m_mode', 'lvid_d_depth_sax_m_mode', 'lvid_s_depth_sax_m_mode', 'lv_vol_d_sax_m_mode', 'lv_vol_s_sax_m_mode', 'av_peak_vel', 'aov_peak_vel_std') # 11 variables, + 1 std for av_peak_vel

```

set up models-
ekg variables

+ tpeak_tend_interval_msec- latency from peak of T-wave to end of T-wave
+ t_amplitude_mv- amplitude of T-wave (ventricular repolarization)
+ qrs_interval_msec- duration of QRS complex (ventricular depolarization)
+ jt_interval_msec- latency from end of QRS complex (S-wave) to end of T-wave (ventricular repolarization)
+ rr_interval_msec- latency between successive R-waves
+ p_duration_msec- duration of P-wave (atrial depolarization)
+ st_amplitude_mv- latency from end of QRS complex (ventricular depolarization) to start of T-wave (ventricular repolarization)
```{r}
# put df together
ekg_model_df <- data_filtered %>% dplyr::select(mouse_id:genotype_manifest, ekg_vars)
# scale each of the variables to put all on the same scale
ekg_model_df_scaled <- ekg_model_df %>%
  mutate(across(all_of(ekg_vars), scale))

# set up multivariate response regression model using all variables as 'one'
strain_ekg_model <- lm(cbind(as.matrix(ekg_model_df_scaled[, ekg_vars]))  ~ jr_strain , data = ekg_model_df_scaled)

summary(strain_ekg_model) # we are using model with strain as the only predictor because that is what we are interested in

```

Does strain have an overall multivariate effect across all ekg phenotypes considered together?
```{r}

anova(strain_ekg_model) # manova since multiple response variables were combined into one

```

```{r}
# marginal means by strain (averaged over sex) for ALL ekg variables
strain_ekg_marginal_means <- ekg_model_df_scaled %>%
  group_by(jr_strain) %>%
  summarise(across(all_of(ekg_vars), ~ mean(.x, na.rm = TRUE)), .groups = 'drop')

print(strain_ekg_marginal_means)

# SE and CI 95% for marginal means
se <- function(x) sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
ci95 <- function(x) {
  n <- sum(!is.na(x))
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  error <- qt(0.975, df = n - 1) * s / sqrt(n)
  c(lower = m - error, upper = m + error)
}

# get stats and format 
strain_ekg_summary <- ekg_model_df_scaled %>%
  group_by(jr_strain) %>%
  summarise(across(all_of(ekg_vars),
                   list(mean = ~ mean(.x, na.rm = TRUE),
                        SE = ~ se(.x),
                        CI_low = ~ ci95(.x)[1],
                        CI_high = ~ ci95(.x)[2]),
                   .names = "{.col}_{.fn}"),
            .groups = "drop")

# format mean ± SE if desired
strain_ekg_formatted <- strain_ekg_summary %>%
  rowwise() %>%
  mutate(across(ends_with("_mean"),
                ~ paste0(round(.x, 2), " ± ",
                         round(cur_data()[[sub("_mean", "_SE", cur_column())]], 2)),
                .names = "{.col}_formatted"))

# print table
print(strain_ekg_formatted)

# marginal means by sex (averaged over strain) for ALL ekg variables  
sex_ekg_marginal_means <- ekg_model_df_scaled %>%
  group_by(sex) %>%
  summarise(across(all_of(ekg_vars), ~ mean(.x, na.rm = TRUE)), .groups = 'drop')

print(sex_ekg_marginal_means)

```

Does strain significantly affect the specific phenotypes?
```{r}

# this pulls the F stat, etc from the model
# values are the same as if you run summary(model), but this is a cleaner way to pull the F statistic, etc.
summary_aov_ekg<- summary.aov(strain_ekg_model)

summary_aov_ekg_results <- purrr::map_dfr(seq_along(summary_aov_ekg), function(i) {
  tab <- summary_aov_ekg[[i]]
  tibble(
    variable = names(summary_aov_ekg)[i],
    df1      = tab["jr_strain", "Df"],
    df2      = tab["Residuals", "Df"],
    F        = tab["jr_strain", "F value"],
    p        = tab["jr_strain", "Pr(>F)"]
  )
})

summary_aov_ekg_results <- summary_aov_ekg_results %>%
  mutate(p_adj = p.adjust(p, method = "BH"))

summary_aov_ekg_results

```

How does strain affect this phenotype in terms of direction and magnitude?
```{r}

# account for multiple testing
# pull values from model summary (estimates, pvalues)
strain_ekg_model_summary <- summary(strain_ekg_model)

# initialize empty vectors to store values
response_names <- names(strain_ekg_model_summary)
all_pvalues <- c()
all_coefficients <- c()
variable_labels <- c()

# go through each response variable
for(i in 1:length(response_names)) {
  # get p-values for jr_strain coefficient
  p_val <- strain_ekg_model_summary[[i]]$coefficients[2, "Pr(>|t|)"]
  print(p_val)
  coef_val <- strain_ekg_model_summary[[i]]$coefficients[2, "Estimate"]
  print(coef_val)
  
  all_pvalues <- c(all_pvalues, p_val)
  all_coefficients <- c(all_coefficients, coef_val)
  variable_labels <- c(variable_labels, paste0(response_names[i], "_jr_strain"))
  
  if(i == length(response_names)){
    break
  }
}

# apply BH correction across all ekg variables
adjusted_pvalues <- p.adjust(all_pvalues, method = "BH")

# create results table
results <- data.frame(
  ekg_variable = response_names,
  coefficient = all_coefficients,
  original_p = all_pvalues,
  adjusted_p = adjusted_pvalues,
  significant_original = all_pvalues < 0.05,
  significant_adjusted = adjusted_pvalues < 0.05
)

# print table
print(results)

```

echo variables
pulled variables thought to be most important with respect to marfan syndrome and cardiovascular health in relation to an echo

+ asc_ao- diameter of ascending aortic valve
+ linear2_aortic_annulus- diameter of aortic annulus
+ linear3_aortic_sinus- diameter of aortic sinus
+ linear4_sinotabular_junction- diameter of sinotubular junction
+ ef_sax_m_mode- ejection fraction (percentage of blood pumped out of heart with each beat)
+ fs_sax_m_mode- fractional shortening (percentage of ventricular size difference before and after ventricular contraction)
+ lvid_d_depth_sax_m_mode- left ventricular interior diameter standard deviation during diastole
+ lvid_s_depth_sax_m_mode- left ventricular interior diameter standard deviation during systole- what about the std variables?
+ lv_vol_d_sax_m_mode- left ventricular volume during diastole
+ lv_vol_s_sax_m_mode- left ventricular volume during systole
+ av_peak_vel- measures the maximum blood flow velocity across the aortic valve- what about the std variable?
+ aov_peak_vel_std- standard deviation

std for av_peak_vel (aov_peak_vel_std)- is large (range 8-99), so using to calculate mean/std to then use that in the model 

```{r}

# put df together
echo_model_df <- data_filtered %>% dplyr::select(mouse_id:genotype_manifest, echo_vars)
# divide av_peak_vel by aov_peak_vel_std
echo_model_df <- echo_model_df %>% mutate(av_peak_vel = av_peak_vel/aov_peak_vel_std) %>% dplyr::select(-aov_peak_vel_std)

# remove aov_peak_vel_std from the echo_vars vector
echo_vars <- echo_vars[echo_vars != "aov_peak_vel_std"]

# scale each of the variables to put all on the same scale
echo_model_df_scaled <- echo_model_df %>%
  mutate(across(all_of(echo_vars), scale))

# set up multivariate response regression model using all variables as 'one'
strain_echo_model <- lm(cbind(as.matrix(echo_model_df_scaled[, echo_vars]))  ~ jr_strain , data = echo_model_df_scaled)

summary(strain_echo_model) # we are using model with strain as the only predictor because that is what we are interested in

```

Does strain have an overall multivariate effect across all echo phenotypes considered together?
```{r}

anova(strain_echo_model)

```

```{r}

# marginal means by strain (averaged over sex) for ALL echo variables
strain_echo_marginal_means <- echo_model_df_scaled %>%
  group_by(jr_strain) %>%
  summarise(across(all_of(echo_vars), ~ mean(.x, na.rm = TRUE)), .groups = 'drop')

print(strain_echo_marginal_means)

# SE and CI 95% for marginal means
se <- function(x) sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
ci95 <- function(x) {
  n <- sum(!is.na(x))
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  error <- qt(0.975, df = n - 1) * s / sqrt(n)
  c(lower = m - error, upper = m + error)
}

# get stats and format
strain_echo_summary <- echo_model_df_scaled %>%
  group_by(jr_strain) %>%
  summarise(across(all_of(echo_vars),
                   list(mean = ~ mean(.x, na.rm = TRUE),
                        SE = ~ se(.x),
                        CI_low = ~ ci95(.x)[1],
                        CI_high = ~ ci95(.x)[2]),
                   .names = "{.col}_{.fn}"),
            .groups = "drop")

# format mean ± SE if desired
strain_echo_formatted <- strain_echo_summary %>%
  rowwise() %>%
  mutate(across(ends_with("_mean"),
                ~ paste0(round(.x, 2), " ± ",
                         round(cur_data()[[sub("_mean", "_SE", cur_column())]], 2)),
                .names = "{.col}_formatted"))

# print table
print(strain_echo_formatted)

# marginal means by sex (averaged over strain) for ALL echo variables  
sex_echo_marginal_means <- echo_model_df_scaled %>%
  group_by(sex) %>%
  summarise(across(all_of(echo_vars), ~ mean(.x, na.rm = TRUE)), .groups = 'drop')

print(sex_echo_marginal_means)

```

Does strain significantly affect the specific phenotypes?
```{r}

# this pulls the F stat, etc from the model
# values are the same as if you run summary(model), but this is a cleaner way to pull the F statistic, etc.
summary_aov_echo <- summary.aov(strain_echo_model)

summary_aov_echo_results <- purrr::map_dfr(seq_along(summary_aov_echo), function(i) {
  tab <- summary_aov_echo[[i]]
  tibble(
    variable = names(summary_aov_echo)[i],
    df1      = tab["jr_strain", "Df"],
    df2      = tab["Residuals", "Df"],
    F        = tab["jr_strain", "F value"],
    p        = tab["jr_strain", "Pr(>F)"]
  )
})

summary_aov_echo_results <- summary_aov_echo_results %>%
  mutate(p_adj = p.adjust(p, method = "BH"))

summary_aov_echo_results

```

How does strain affect this phenotype in terms of direction and magnitude?
```{r}

#account for multiple testing
# pull values from model summary (estimates, pvalues)
strain_echo_model_summary <- summary(strain_echo_model)

# initialize empty vectors to store values
response_names <- names(strain_echo_model_summary)
all_pvalues <- c()
all_coefficients <- c()
variable_labels <- c()

# go through each response variable
for(i in 1:length(response_names)) {
  # get p-values for jr_strain coefficient
  p_val <- strain_echo_model_summary[[i]]$coefficients[2, "Pr(>|t|)"]
  coef_val <- strain_echo_model_summary[[i]]$coefficients[2, "Estimate"]
  
  all_pvalues <- c(all_pvalues, p_val)
  all_coefficients <- c(all_coefficients, coef_val)
  variable_labels <- c(variable_labels, paste0(response_names[i], "_jr_strain"))
  
   if(i == length(response_names)){
    break
   }
  
}

# apply BH correction across all echo variables
adjusted_pvalues <- p.adjust(all_pvalues, method = "BH")

# create results table
results <- data.frame(
  echo_variable = response_names,
  coefficient = all_coefficients,
  original_p = all_pvalues,
  adjusted_p = adjusted_pvalues,
  significant_original = all_pvalues < 0.05,
  significant_adjusted = adjusted_pvalues < 0.05
)

print(results)

```


