---
title: "Strain Recommender Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Strain Recommender Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(recommender)
```

## Overview

Strain Recommender is a computational tool that identifies which inbred 
mouse strains are most biologically relevant to a disease of interest. Rather 
than relying solely on traditional phenotypic similarity, it uses transcriptomic 
data to match a disease-associated gene expression signature against predicted 
gene expression profiles for 657 inbred mouse strains, returning a vulnerability 
score and false positive rate (FPR) for each strain.

The main inputs are:

- **`gs`** — a data frame of gene scores representing a disease-associated gene 
  expression signature, with columns `gene_id` (mouse Ensembl IDs) and `score`
- **`meta`** — a data frame of sample metadata with columns `mouse_id`, `sex`, 
  and `strain`
- **`df`** — a data frame of gene expression data with `gene_id` as a column 
  and samples as remaining columns

---

## Preparing Input Data

Users will need to read their data in (gs, meta, and df) before passing 
them to `recommend()`. For example:

```{r eval=FALSE}
gs   <- fread("path/to/gene_scores.csv", data.table = FALSE)
meta <- fread("path/to/meta.csv", data.table = FALSE)
df   <- fread("path/to/df.csv", data.table = FALSE)
```

Each input must be a data frame. See the `?recommend` documentation for full 
column requirements.

---

## Simulated Example Data

To illustrate use of the package, we use simulated data.

```{r}
set.seed(42)

# Gene scores: disease-associated signature
gs <- data.frame(
  gene_id = c("ENSMUSG00000000168", "ENSMUSG00000000544", "ENSMUSG00000000792",
              "ENSMUSG00000001017", "ENSMUSG00000001240", "ENSMUSG00000003068",
              "ENSMUSG00000003134", "ENSMUSG00000003464", "ENSMUSG00000003644",
              "ENSMUSG00000004768"),
  score = c(1.5, -2.3, 0.8, -1.1, 2.0, -0.5, 1.7, -1.9, 0.6, 2.2)
)

# Metadata: one sample per strain
meta <- data.frame(
  mouse_id = paste0("sample", 1:10),
  sex      = rep("both", 10),
  strain   = paste0("strain", LETTERS[1:10])
)

# Expression data: genes x samples
df <- data.frame(
  gene_id = gs$gene_id,
  matrix(rnorm(100, mean = 5, sd = 1.5), nrow = 10, ncol = 10,
         dimnames = list(NULL, paste0("sample", 1:10)))
)
```

---

## Basic Workflow: Single Gene Set

The simplest use case is a single gene set passed directly to `recommend()`:

```{r}
result <- recommend(gs   = gs,
                    meta = meta,
                    df   = df,
                    B    = 100,       # use B=4000 in practice
                    verbose = FALSE)

head(result)
```

The output is a data frame with one row per strain containing:

- **`strain`** — inbred mouse strain name
- **`score`** — scaled Spearman correlation (vulnerability score) between the strain's 
  predicted expression and the disease gene signature for easier interpretation.
  Higher positive values indicate greater biological similarity to the disease state.
- **`corr`** — Spearman correlation (vulnerability score) between the strain's 
  predicted expression and the disease gene signature. Higher positive values 
  indicate greater biological similarity to the disease state.
- **`n`** — number of genes used in the prediction
- **`fpr`** — false positive rate, estimated from a bootstrapped null 
  distribution. Lower FPR indicates greater confidence in the vulnerability score.

---

## Interpreting the Output

Sort strains by vulnerability score to identify top candidates:

```{r}
result_sorted <- result[order(result$score, decreasing = TRUE), ]
head(result_sorted)
```

Strains with high positive `corr` and low `fpr` are the strongest candidates — 
they are most transcriptomically similar to the disease-associated gene signature.

A general interpretation guide:

| Score Range     | Interpretation          | Predicted Phenotype                    | Recommended Use                   |
|-----------------|-------------------------|----------------------------------------|-----------------------------------|
| +0.5 to +1.0   | High Vulnerability      | Strongly mimics disease expression     | Disease model development         |
| +0.1 to +0.5   | Moderate Vulnerability  | Partial alignment with disease         | May be suitable; check confidence |
| −0.1 to +0.1   | Indeterminate           | Weak / no alignment                    | Consider alternative strains      |
| −0.5 to −0.1   | Moderate Resistance     | Partial inverse of disease             | Control cohort candidate          |
| −1.0 to −0.5   | High Resistance         | Strongly inversely aligned             | Protective / resilience studies   |

---

## Advanced Workflow: Multiple Gene Sets

If your gene scores come from multiple sources or studies, you can normalize 
them onto a common scale using `combine_genesets()` before running `recommend()`.
This requires a `geneset` column in `gs` to distinguish which genes comes from which genes set:

```{r}
gs_multi <- data.frame(
  gene_id = c("ENSMUSG00000000168", "ENSMUSG00000000544", "ENSMUSG00000000792",
              "ENSMUSG00000001017", "ENSMUSG00000001240", "ENSMUSG00000003068",
              "ENSMUSG00000003134", "ENSMUSG00000003464", "ENSMUSG00000003644",
              "ENSMUSG00000004768"),
  score   = c(1.5, -2.3, 0.8, -1.1, 2.0, -0.5, 1.7, -1.9, 0.6, 2.2),
  geneset = c(rep("study1", 5), rep("study2", 5))
)
```

### Option 1: Inspect combined gene set first

You can preview the normalized gene scores before passing to `recommend()`:

```{r}
gs_combined <- combine_genesets(gs_multi)
head(gs_combined)
```

Then pass the combined gene set to `recommend()`:

```{r}
result_combined <- recommend(gs      = gs_combined,
                             meta    = meta,
                             df      = df,
                             B       = 100,
                             combine_gs = FALSE, # this is the deafult.
                             verbose = FALSE)
head(result_combined)
```

Note: `combine_gs` default is FALSE. It should be set to FALSE when using a singular gene set or one that has already been combined.

### Option 2: Combine within `recommend()`

Alternatively, pass the multi-geneset data frame directly to `recommend()` with 
`combine_gs = TRUE`:

```{r}
result_combined2 <- recommend(gs      = gs_multi,
                              meta    = meta,
                              df      = df,
                              B       = 100,
                              combine_gs = TRUE,
                              verbose = FALSE)
head(result_combined2)
```

Both options produce identical results. Option 1 is useful when you want to 
inspect or save the normalized gene scores before proceeding.

---

## Reproducibility

The FPR calculation relies on a bootstrapped null distribution. Use the `seed` 
argument to ensure reproducibility:

```{r eval=FALSE}
result <- recommend(gs      = gs,
                    meta    = meta,
                    df      = df,
                    seed    = 303,
                    B       = 4000,
                    verbose = FALSE)
```

---

## Session Info

```{r}
sessionInfo()
```
